### 2016-06-28 publications/Journal/4a-allocation.tex

%TODO explain model allocation
%TODO why model allocation
Modeller can specify which kernel a model should be allocated to, should such manual intervention be required. This is handled by the default model allocator. If no preference is given a simple striping scheme is followed but this is not sufficient in most simulations to achieve a speedup in parallel. In Section \nameref{sec:4-performance} we analyse the performance impact of allocation schemes. By overriding the default allocator a model author can tune the allocation scheme for a specific model to maximize parallel speedup. This interface can be in future work also be used to employ graph algorithms for an automatic allocation scheme, for example one that avoids cycles in the resulting kernel dependency graph.

\subsection{Performance Evaluation}
%TODO do basic performance evaluation of allocation
%TODO show model in statistics view

The \textit{PHoldTree} model, a hierarchical model based on the semantics of PHOLD~\cite{PHOLD}. The model consists of recursively defined trees connected to each other, similar to the structure of an AST. Connections can be unidirectional or bidirectional.
Unlike the \textit{Queue} model the width of the hierarchy is still present in the topology of the atomic models after the directconnect stage. This model allows us to investigate parallel speedup in several key dimensions in combination or isolation : model allocation, message density, uncertainty and hierarchy. The model has the following parameters : depth, fanout (width), probability of priority message and linkage direction. We demonstrate the importance of atomic to kernel allocation, in this model this can be configured as a bread-first versus a depth-first scheme.
\textit{PHOLDTree} can be used to model gossiping in social networks, hierarchical resource sharing or load balancing. Unlike \textit{Queue} this model does not lend itself easily to parallelism and thus highlights the difficulties model authors face in when a parallel speedup is desired for real world models.
The lookahead of an atomic node is $\epsilon$, simulating uncertainty as will be often the case in real world models.

\begin{figure}
\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtree.pdf}
\caption{PHOLDTree model for depth 1, width 2.}
\label{fig:PHOLDTree_model}
\end{figure}

>>>>

Although the synchronization protocol is one of the defining factors in simulation performance, model allocation has a significant impact on which protocol is ideal, and even whether or not parallelization would make sense.
Indeed, if the model is distributed in such a way that frequent communication is necessary between cores, parallelism is naturally reduced.
This thus brings us to the topic of model allocation.
Model allocation was one of the features also implemented by PythonPDEVS~\cite{PythonPDEVS2}.

The modeller can specify which kernel a model should be allocated to, should such manual intervention be required.
This is handled by the default model allocator.
If no preference is given a simple striping scheme is followed but this is not sufficient in most simulations to achieve a speedup in parallel.
By overriding the default allocator a modeller tunes the allocation scheme for a specific model, maximizing parallel speedup.
This interface can be used to employ graph algorithms for an automatic allocation scheme, for example avoiding cycles in the dependency graph.

\subsection{Performance Evaluation}
%TODO do basic performance evaluation of allocation
%TODO show model in statistics view
In order to evaluate the influence of model allocation, we define a new model, based on PHOLD~\cite{PHOLD}.
The model structure resembles a tree: an atomic model can have a set of children, with children being connected to each other as well.
Connections can be uni- or bidirectional.

Unlike the Queue model, the width of the hierarchy is still present in the topology of the atomic models after the direct connection stage.
The PHOLDTree model allows us to investigate parallel speedup in terms of model allocation, by modifying the depth and width (fanout) model parameters.

PHOLDTree can be used to model gossiping in social networks, hierarchical resource sharing or load balancing.
The lookahead of an atomic node is $\epsilon$, simulating uncertainty as will be often the case in real world models.
We demonstrate the importance of allocation, by comparing a breadth-first versus a depth-first scheme.

\begin{figure}
\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtree.pdf}
%TODO depth 1 but 3 levels deep?
\caption{PHOLDTree model for depth 1 and width 2.}
\label{fig:PHOLDTree_model}
\end{figure}

### 2016-06-28 publications/Journal/conclusion.tex
New order of sections

\subsection{Conclusion}
We have shown that our contribution is invaluable for high performance simulation: depending on the expected behaviour, modellers can choose the most appropriate synchronization protocol.
We will summarize the key factors in achieving good parallel performance here. % reword aspects syn
\subsubsection{Allocation}
Allocation of atomic models over kernels is critical to obtain a speedup. 
In the case of the Interconnect model it is not possible to allocate models without introducing a cyclic dependency between kernels with severe performance degradation as a result. 
Even when no cycles exist in the topology between kernels a good allocation scheme will minimize the number of connections between kernels, for example the star topology in PHoldTree with depth first allocation or the chain topology in the Queue model. 
Dxex can optionally offer the user a visualization of the behaviour of the kernels under a user supplied allocation scheme to allow for more insight. 
The same instrumentation could in future be used to optimize an existing allocation scheme.
\subsubsection{Uncertainty}
A simulation where future behaviour cannot be predicted is typically not suited for conservative simulation, but we have demonstrated that dxex's conservative kernel can still offer a non trivial speedup in this context. 
Optimistic is not limited by this uncertainty and can offer a good speedup regardless of lookahead. 
Phold and PholdTree demonstrate that a single model can benefit from different synchronization protocols depending on a single parameter.
\subsubsection{Limits}
Dxex's conservative kernel is constrained by the amount of models it controls, as shown in the PHoldTree benchmark. This is especially the case when the minimal lookahead is $\epsilon$ requiring near constant polling of models for lookahead values. 
Optimistic is not constrained by the number of models it manages, but can quickly consume all available memory in a simulation. While this effect can be reduced with a thread aware allocation scheme, the underlying problem remains.

>>>>

\subsection{Conclusions on Performance Evaluation}
We have shown that our contribution is invaluable for high performance simulation: depending on the expected behaviour, modellers can choose the most appropriate synchronization protocol.
But even with the right synchronization protocol, we have seen that two problems remain.

First, although one of both synchronization protocols might be ideally suited for specific model behaviour, nothing guarantees that the model will exhibit the same behaviour throughout the simulation.
Similarly to the polymorphic scheduler~\cite{MasterThesis}, we wish to make it possible for the ideal option to be switched during simulation.
When changes to the model behaviour are noticed, the used synchronization protocol is modified as well.

Second, the allocation of models is tricky and has a significant impact on performance.
While our parallel speedup for the Queue model, for example, was rather high, this is mostly due to characteristics of the model: the dependency graph does not contain any cycles.
When cycles were introduced, as in the Interconnect model, performance became disastrous.

In the next two sections, we elaborate on these two problems.

### 2016-06-28 publications/Journal/parallel-performance.tex

\paragraph*{Strong and Weak Scaling}
Figure~\ref{fig:Queue_plot_strong} shows the speedup compared to dxex sequential for a fixed problem size. As the amount of kernels increases, the optimistic kernel quickly becomes the worst choice. The difference between dxex conservative and adevs becoming smaller. The same effect can be seen for weak scaling in Figure~\ref{fig:Queue_plot_weak}.\\
In Figure \ref{fig:Queue_allocation} the allocation of the Queue model is visualized. This trace allows us to demonstrate the benchmark results and explain why optimistic is not ideal for such a chain topology as present in Queue. Kernel 2 is dependent on events from kernels 0 and 1 but by definition of the optimistic synchronization protocol it will simulate ahead without waiting for events it depends on. Any simulation it performs will have to be reverted as soon as events from kernels 0 and 1 arrive. When kernel 2 reverts, kernel 3 will inevitably have to revert as well since most of the events it received from kernel 2 are invalidated and marked as such by antimessages. This leads to severe performance degradation, with an increasing probability of cascading reverts as the number of kernels and models per kernels increases.
The speedup of adevs is always in comparison with the runtime of the corresponding dxex sequential benchmark.
\begin{figure}
	\center

@@

\begin{figure}
	\center

	\includegraphics[width=\modelfraction\columnwidth, height=8cm, keepaspectratio, angle=-90 ]{fig/queue_allocation.eps}
	\caption{Queue model (d=2, w=7, t=5000, random timeadvance) allocation and simulation trace across 4 kernels.}
	\label{fig:Queue_allocation}

@@

\begin{figure}
    \center
    \includegraphics[width=\plotfraction\columnwidth]{fig/phold_remotes.eps}
    \caption{Phold benchmark results for parallel simulation using four kernels, four atomics per node, with varying percentage of remote events.}
\end{figure}
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/phold_priority.eps}
	\caption{Phold benchmark results for parallel simulation using four kernels, with varying amount of high-priority events.}
	\label{fig:phold_priority}
\end{figure}
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/phold_parallel_allocation.eps}
	\caption{Phold benchmark trace for parallel simulation using three kernels.}
	\label{fig:phold_allocation}
\end{figure}

>>>>

\paragraph*{Strong and Weak Scaling}
Figure~\ref{fig:Queue_plot_strong} shows the speedup compared to dxex sequential for a fixed problem size.
As the amount of kernels increases, the optimistic kernel quickly becomes the worst choice.
The difference between dxex conservative and adevs becoming smaller.
The same effect can be seen for weak scaling in Figure~\ref{fig:Queue_plot_weak}.

%In Figure \ref{fig:Queue_allocation} the allocation of the Queue model is visualized. This trace allows us to demonstrate the benchmark results and explain why optimistic is not ideal for such a chain topology as present in Queue. Kernel 2 is dependent on events from kernels 0 and 1 but by definition of the optimistic synchronization protocol it will simulate ahead without waiting for events it depends on. Any simulation it performs will have to be reverted as soon as events from kernels 0 and 1 arrive. When kernel 2 reverts, kernel 3 will inevitably have to revert as well since most of the events it received from kernel 2 are invalidated and marked as such by antimessages. This leads to severe performance degradation, with an increasing probability of cascading reverts as the number of kernels and models per kernels increases.
The speedup of adevs is always in comparison with the runtime of the corresponding dxex sequential benchmark.
\begin{figure}
	\center

@@

@@

\begin{figure}
    \center
    \includegraphics[width=\plotfraction\columnwidth]{fig/phold_remotes.eps}
    \caption{Phold benchmark results for parallel simulation using four kernels, four atomics per node, with varying percentage of remote events.}
\end{figure}
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/phold_priority.eps}
	\caption{Phold benchmark results for parallel simulation using four kernels, with varying amount of high-priority events.}
	\label{fig:phold_priority}
\end{figure}

### 2016-06-28 publications/Journal/sequential-performance.tex

\subsection{Sequential Simulation}
% 2 things to show : baseline performance and contrast of dxex v adevs
\subsubsection{Queue}\label{4-seq-Queue}
% explain what we did: fixed number of models (400, 600 & 800), increasing depth
For the first benchmark, we tested the effect of hierarchical complexity of the model in the performance of the simulator. A set of three tests was performed, where each test has the same number of models but an increasing depth.
% compare dxex & adevs performance
The results can be seen in Figure~\ref{fig:Queue_benchmark_seq}. Since dxex symbolically flattens the model, there is no performance hit when the depth is increased. The overhead of running the directconnect algorithm is one time only and negligible when the end time of the simulation is sufficiently large. Adevs on the other hand does suffer from the increased depth. With every new hierarchical layer, routing an event from one atomic model to the next becomes more expensive, resulting in an increase in runtime.
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/queue_fixed_sequential.eps}
	\caption{Queue benchmark results for sequential simulation.}
	\label{fig:Queue_benchmark_seq}
\end{figure}
\subsubsection{Interconnect}\label{4-seq-Interconnect}
In the Interconnect model, we increase the number of atomic models, thus quadratically increasing the number of couplings and the number of external transitions.
As can be seen in Figure~\ref{fig:Interconnect_benchmark}, adevs outperforms dxex by a fair margin.
Analysis showed that this is caused by the high amount of events: event creation is much slower in dxex than it is in adevs, despite dxex's use of memory pools. To shield the user from threading and deallocation concerns dxex provides an event superclass from which the user can derive to create a specialized event type. Copying and deallocation semantics and tracing are solved by the kernels at a runtime cost in simulations where event frequency is very high. Profiling the benchmarks clearly shows the increasing cost of output generation and deallocation as the determining factor in the gap in performance. We refer the interested reader to the dxex \hyperref{https://bitbucket.org/bcardoen/devs-ex-machina}{}{repo}{repository} for the profiling call graphs for the different benchmarks.

\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/interconnect_sequential.eps}
	\caption{Interconnect benchmark results for sequential simulation.}
	\label{fig:Interconnect_benchmark}
\end{figure}
% We want to show the effect of message storms, that dxex is sensitive to this (even with allocators)
\subsubsection{PHold}\label{4-seq-PHold}
% Phold : difference with interconnect, less messages == equal performance dxex adevs
The PHold model is very similar to the Interconnect model. The biggest difference is that the amount of messages sent is much lower. The number of events scales linear with the number of models, not quadratic. Figure~\ref{fig:Phold_benchmark} shows that the performance of dxex and adevs are very close to each-other, with adevs slightly outperforming dxex.
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/phold_sequential.eps}
	\caption{PHold benchmark results for sequential simulation.}
	\label{fig:Phold_benchmark}
\end{figure}
\subsubsection{PHoldTree}
PHoldtree, like Queue, is a highly hierarchical model but one where the flattened structure cannot be partitioned into a chain as in Queue.
This topology is interesting since it highlights the effects of allocation which we have shown already in PHold and Interconnect to be of vital importance. This model allows us to investigate in depth the effects of non-cyclic allocation strategies and measure parallel speedup.\\
Since adevs does not use the directconnect algorithm, we expect some performance penalty between dxex and adevs. The event frequency in PHoldTree can be configured to highlight dxex's performance disadvantage here as shown in the Interconnect benchmark.
\paragraph*{Event frequency}
Similarly when the probability on a priority event (p) increases the amount of event generated will increase and thus the runtime. In Figure \ref{fig:PHoldtree_seq_p_benchmark} we observe that p has a near linear impact on performance, with the shape of the curve due to rounding errors. The advantage adevs has in simulations where event frequency is high has been demonstrated by the interconnect benchmark. 
\paragraph*{Hierarchy}
To establish a baseline for the parallel simulation benchmarks we measure how the PHoldTree benchmark performs sequentially when we vary each of the model's parameters. 
When we increase the depth and fanout of the model, we expect to see an increase in runtime. In Figure \ref{fig:PHoldtree_seq_dn_benchmark} the n parameter (fanout) determines the performance penalty adevs suffers compared to dxex.\\
If we compare the instances d=2, n=2 with d=2, n=4, which show a clear difference in performance between adevs and dxex, we conclude from the profiling callgraphs that the increase in width per subtree (n) leads to higher overhead for adevs. Dxex in contrast uses the directconnect algorithm and has no such overhead. An increase in n will rapidly increase the number of connections \ref{eq:pholdtreelinkcount} (and thus the routing problem adevs faces), whereas d will increase modelcount \ref{eq:pholdtreemodelcount} faster but have a lesser impact on connections. Both kernels scale linear in increasing number of atomic models. 
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/pholdtree_sequential_p.eps}
	\caption{PHoldTree for increasing probability of priority message.}
	\label{fig:PHoldtree_seq_p_benchmark}
\end{figure}
% Img for p increasing
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/pholdtree_sequential_dn.eps}
	\caption{PHoldTree : Effect of hierarchy.}
	\label{fig:PHoldtree_seq_dn_benchmark}
\end{figure}

>>>>

\subsection{Sequential Simulation}
% 2 things to show : baseline performance and contrast of dxex v adevs

\subsubsection{Queue}
\label{4-seq-Queue}
% explain what we did: fixed number of models (400, 600 & 800), increasing depth
For the first benchmark, we tested the effect of hierarchical complexity of the model in the performance of the simulator.
A set of three tests was performed, where each test has the same number of models but an increasing depth.
The results can be seen in Figure~\ref{fig:Queue_benchmark_seq}.
Since dxex symbolically flattens the model, there is no performance hit when the depth is increased.
The overhead of running the directconnect algorithm is one time only and negligible when the end time of the simulation is sufficiently large.
Adevs on the other hand does suffer from the increased depth.
With every new hierarchical layer, routing an event from one atomic model to the next becomes more expensive, resulting in an increase in runtime.
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/queue_fixed_sequential.eps}
	\caption{Queue benchmark results for sequential simulation.}
	\label{fig:Queue_benchmark_seq}
\end{figure}

\subsubsection{Interconnect}
\label{4-seq-Interconnect}
In the Interconnect model, we increase the number of atomic models, thus quadratically increasing the number of couplings and the number of external transitions.
As can be seen in Figure~\ref{fig:Interconnect_benchmark}, adevs outperforms dxex by a fair margin.
Analysis showed that this is caused by the high amount of events: event creation is much slower in dxex than it is in adevs, despite dxex's use of memory pools.
To shield the user from threading and deallocation concerns dxex provides an event superclass from which the user can derive to create a specialized event type.
Copying and deallocation semantics and tracing are solved by the kernels at a runtime cost in simulations where event frequency is very high.
Profiling the benchmarks clearly shows the increasing cost of output generation and deallocation as the determining factor in the gap in performance.
We refer the interested reader to the dxex \hyperref{https://bitbucket.org/bcardoen/devs-ex-machina}{}{repo}{repository} for the profiling call graphs for the different benchmarks.

\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/interconnect_sequential.eps}
	\caption{Interconnect benchmark results for sequential simulation.}
	\label{fig:Interconnect_benchmark}
\end{figure}

\subsubsection{PHold}
\label{4-seq-PHold}
The PHold model is very similar to the Interconnect model.
The biggest difference is that the amount of messages sent is much lower.
The number of events scales linear with the number of models, not quadratic.
Figure~\ref{fig:Phold_benchmark} shows that the performance of dxex and adevs are very close to each-other, with adevs slightly outperforming dxex.
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth]{fig/phold_sequential.eps}
	\caption{PHold benchmark results for sequential simulation.}
	\label{fig:Phold_benchmark}
\end{figure}
